Substitution rate responses to changes in effective population size.

### 2020-05-09 manuscript/manuscript.tex
Added Figures and updated manuscript

\documentclass{article}
\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1
\captionsetup{width=0.85\textwidth}
\include{notations}
\begin{document}
\title{Substitution rate responses to changes in effective population size.}
\author{T. Latrille, N. Lartillot}
\maketitle

\abstract{
The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
Such theoretical prediction had been observed in empirical data across many clades.
However, several studies failed to observe such response of $\dnds$ to changes in $\Ne$, or with weak strength or direction.
Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
We apply our framework in the special case of fitness proportional to the probability of protein folding.
Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
}

@@

Differences in DNA sequences of a gene in separate species are due to the particular history of DNA substitutions along each specie's respective lineage.
These substitutions at the level of the population are the result of point mutations at the level of individuals, subject to evolutionary forces such as selection and random drift.
In protein-coding DNA sequences, non-synonymous mutations are changing the amino-acid sequence, and by comparing the non-synonymous substitutions rate relative to the synonymous substitution rate (supposedly neutral), a statistic called $\dnds$, we can estimate the global strength of selection and random drift.\\
For a given gene, it has been empirically observed changes of $\dnds$ along lineages of a phylogeny \cite{Yang2001, Zhang2004}.
More importantly, changes in substitution rate correlate with changes in life-history-traits (longevity, body mass, ...) such as shown by the molecular comparative framework \cite{Lartillot2011,Weber2014}.
Such covariations are compatible with the theoretical prediction of the nearly neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
Indeed, populations with low $\Ne$ would have both a large body size and a high $\dnds$ (weaker selection) due to the increase of random drift.
If $\dnds$ is empirically correlated to life-history traits, such as body mass and longevity, the direction of correlation could not be replicated across all experiments \cite{Figuet2016}.\\
Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects of novel mutations \cite{Ohta1972, Welch2008}. Similarly, it has been shown in context of a fixed fitness landscape, where each amino-acid have different fitness \cite{Spielman2015a, DosReis2015}, although sites are considered independently evolving, without epistasis.
Unexpectedly, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly neutral theory of evolution \cite{Lanfear2014}.
In the example of a biochemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ as been observed to be independent of $\Ne$ \cite{Goldstein2013}.
Theoretically, a more general case is obtained whenever the fitness is a concave function of a trait, and such trait is equimutable \cite{Cherry1998}.
Equimutabibilty meaning a mutation has the same effect on the trait whatever the current trait of the individual.
In addition to this complexity, $\dnds$ response to $\Ne$ can be non-monotonic due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$ (adaptive selection), and subsequently a decrease in $\dnds$ (purifying selection) \cite{Jones2016}.\\
Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
This question is important if we want to extract changes in effective population size from substitution history, as well as to determine the fitness landscape under which protein-coding DNA sequences are evolving.
We provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
We apply our framework in the special case of fitness proportional to the probability of protein folding.
Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.

@@

\subsection*{Theoretical approximations}
In this section we theoretically approximate the change in $\dnds$ at equilibrium after a change in $\Ne$, under a model in which the fitness depends on the genotype of the whole sequence.
Firstly, we decompose the map from genotype to fitness through an intermediary bounded phenotype $0 \leq \x \leq 1$, the proportion of unstable sites in the sequence.
Each site of the sequence ($\Nsite$ sites) can be in one of $\Nstate \geq 2$ states, where only $1$ of those states is assumed to be a stable state, and $\Nstate - 1$ states are unstable.
The optimal phenotype is obtained when all of the $\Nsite$ sites of the sequence are stable ($\x = 0$).\\
To complete the model, the phenotype-fitness map is first considered generally, where the Wrightian fitness of phenotype $\x$ is $f(\x)$.
After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
Most importantly the probabilities to increase or decrease $\x$ are different, and depends on the current phenotype $\x$.
Mutation increasing $\x$ are selected against, while mutations decreasing $\x$ are positively selected. 
Mutation and selection cancel each other at equilibrium, where the expectation of fitness effects of mutations reaching fixation in the population is $0$.
Altogether, the equilibrium phenotype obtained for such mutation-selection balance is denoted $\x\eq$, and more importantly is given by the equation (See Supplementary Materials):
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
\end{align}
This equation of equilibrium phenotype can be further studied to derive the response of $\x\eq$ after a change in $\Ne$, using differential calculus.
Furthemore, assuming that the number of states is large (typically $20$ for amino-acid), we obtain the change in $\omega\eq$ after a change in $\Ne$ as: 
\begin{align}
\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
\end{align}
In other words, the slope of the $\dnds$ elasticity to changes in $\Ne$ (in log scale) is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype. To note, this elasticity is strictly negative for decreasing log-concave fitness functions.\\
So far, our result did not assume any particular parametrization of the phenotype-fitness map.
As an example, already proposed in the literature \cite{Goldstein2011}, fitness is proportional to the probability of our protein to be in the folded state, given by Fermi Dirac distribution: 
\begin{equation}
f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}, 
\end{equation}
where $\deltaGMin < 0$ and $\gamma > 0$ are in kcal/mol, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature. In other words, $\deltaGMin$ is the difference in free energy between folded and unfolded state when all sites are stable. $\gamma$ is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
With such fitness function, the equilibrium phenotype $\x\eq$ is given by the equation: 
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
\label{eq:equilibrium}
\end{align}
The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
\begin{equation}
\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma};
\end{equation}
which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
Moreover, only the free compound parameter $\gamma \Nsite$ has an impact on the slope, an intuition for such result is given in figure \ref{fig:NeChangeInfluence}.
Meaning that under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is approximately $0.002$.
\begin{figure*}[htb!]
 \begin{mdframed}
  \centering
  \includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
  \caption{
   \textbf{Theoretical model.}
   Panel A. Illustration of the Genotype-phenotype-fitness map. The phenotype ($\x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of unstable sites in the sequence. The fitness is a decreasing log-concave function of the phenotype.
   Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$.
   For the special case $K=2$, the equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$. The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate. Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$ (x-axis of the crossing with the blue solid line). On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably. Moreover, changes in $\x\eq$ reflects the changes in $\dnds\eq$ since both are related by a monotonous function (See Supplementary Materials). The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
  }
  \label{fig:NeChangeInfluence}
 \end{mdframed}
\end{figure*}

\subsection*{Simulations confirmation}
Our compact theoretical results are based on assumptions that are broken in practice, for example assuming that 
\begin{enumerate}
 \item $\Nsite$ is large;
 \item selection coefficient is well approximated by the fitness derivative;
 \item genetic code is not taken into account;
 \item all unstable states are equivalently unstable.
\end{enumerate}
We thus first tested the soundness of our theoretical results when such assumptions are relaxed, by simulating DNA sequences evolution (See Methods). 
Firstly, simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}).
Secondly the parameter $\deltaGMin$ hardly changes the slope of the linear regression, as expected also theoretically.
Decreasing $\deltaGMin$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations. Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral. However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.\\

Finally, we compare our results to more complex models of sequence evolution leveraging the $3$D structure of protein \cite{Williams2006, Goldstein2011, Pollock2012}.
In such models, the free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$.
Using extensive simulations, we argue instead that $\dnds$ is approximately linear with log-$\Ne$ (see Figure \ref{fig:GoldsteinVsToy}, panel C).
Moreover, the observed slope match the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
\begin{figure*}[htb!]
\begin{mdframed}
\centering
\includegraphics[width=0.9\textwidth] {artworks/Elasticity.png}
 \caption{
  \textbf{$\bm{\dnds}$ elasticity to change in $\bm{\Ne}$}.
  For each population size, $100$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
  Panel A. The fixed parameters are $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid. $\alpha$ are given in the legend. Decreasing $\alpha$ (to more negative values) increases $\dnds$ but changes hardly the slope of the linear regression, as expected theoretically.
  $\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
  Panel B. The fixed parameters are $\Nsite=300$ and $\beta=1.686$. Parameters $\alpha$ and $\gamma$ are given in the legend.
  $\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
  The slope of $\dnds$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
  Panel C. In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally \cite{Goldstein2013}.
  This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.

>>>>

\documentclass{article}
\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1
\captionsetup{width=0.85\textwidth}
\renewcommand{\baselinestretch}{1.5}

\include{notations}
\begin{document}
	\title{Substitution rate responses to changes in effective population size.}
	\author{T. Latrille, N. Lartillot}
	\maketitle

	\abstract{
		The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
		One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
		In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$) with respect to $\Ne$.
		Such theoretical prediction had been observed in empirical data across many clades.
		However, several studies failed to observe such response of $\omega$ to changes in $\Ne$, or with weak strength or direction.
		Computational models of protein folding have observed that $\omega$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
		Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\omega$ and then a decrease.
		Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\omega$ responses to changes in $\Ne$.
		Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\omega$ in the context of a genotype-phenotype-fitness map.
		We derive general formulas, and specify an application in the case of fitness proportional to the probability of protein folding.
		Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
		We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\omega$ relationship in population genetic dataset.
		However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
		We also stress the importance of epistatic interactions in the $\Ne$-$\omega$ relationship, such that it determines to time to reach a new equilibrium.
	}

@@

	Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
	These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
	These forces have effects at different levels: mutations are carried by molecular sequence, selection is mediated at the level of individuals, while random genetic drift is a population effect.
	Yet, they jointly contribute to the long-term molecular evolutionary process.
	Thus, the challenge of molecular evolution is to tease out their respective contributions, based on comparative analyses.

	One main aspect of this challenge is to correctly evaluate the role of random drift in the long term evolutionary process.
	Population genetics theory implies that the strength of drift, due to the stochastic sampling of mutations, is less pronounced in lineage with large effective population size ($\Ne$), and as a consequence, the purification by selection of weakly deleterious mutations is more effective.
	This fundamental idea is at the core of the nearly-neutral theory of evolution.
	The nearly-neutral theory posits that a substantial fraction of mutations are weakly deleterious.
	As a result, the theory predicts that the substitution rate of selected mutations relative to the neutral substitution rate, called $\omega$, decreases along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.

	This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects (DFE) \cite{Kimura1979, Welch2008}.
	Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $\omega$ as a function of $\Ne$ (i.e. $\omega \sim \Ne^{-\beta}$), where $\beta$ is the shape parameter of the DFE.
	In practice, DFEs are strongly leptokurtic, thus predicts a weak negative relation between $\omega$ and $\Ne$.

	The substitution rate of selected mutations relative to neutral substitution rate ($\omega$) is an observable quantity of the molecular evolutionary process, at least if we can separately estimate the rate of neutral and selected mutations in a phylogenetic or comparative context.
	Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while those selected mutations are the non-synonymous changes affecting the amino-acid sequence, $\omega$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
	Thus, in practice, the nearly-neutral argument translates into a predicted decrease in $\omega = dN/dS$ as a function of long-term changes in $\Ne$.

	The context of protein coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
	In this alternative approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998, Rodrigue2010, Tamuri2012}.
	For example, if the target amino-acid has a higher fitness than the current amino-acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
	Such a modeling approach predicts higher substitution rates between amino-acid with similar biochemical properties.
	More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
	Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\omega$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.

	Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \cite{Yang2001, Zhang2004}. 
	Combining estimations of $\omega$ along branches to life-history traits such as body mass or generation time revealed a positive correlation \cite{Popadin2007, Nikolaev2007}
	Subsequently, integrative inference methods combining molecular sequences and lineage specific quantitative traits have also found that $\omega$ correlates positively with traits such as longevity and body mass \cite{Lartillot2011, Figuet2017}.
	Since longevity and body mass typically correlate negatively with $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}, these empirical correlations suggest a negative correlation between $\omega$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
	However, this correlation between $\omega$ and life-history traits could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation was found to be either not statistically significant (\cite{Lartillot2012}), or even in the opposite direction \cite{Lanfear2010, Nabholz2013, Weber2014, Figuet2016}.

	Mitigated empirical evidence about the response of $\omega$ to changes in $\Ne$ encouraged the search for alternative explanatory mechanisms \cite{Lanfear2014}.
	In this direction, one striking result was the proof that $\omega$ is in fact predicted to be independent of $\Ne$ under very general circumstances, namely, whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
	Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
	This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \cite{Goldstein2013}.
	The thermodynamic stability is itself computed using a 3D structural model of the protein. These simulation experiments have led to the observation that $\omega$ is essentially independent of $\Ne$.
	The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.

	However, the equimutability assumption is a relatively strong one, which also conflicts with simple combinatorial considerations about the relation between sequence and phenotype. For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur. More generally, assuming that the stability of a protein sequence reflects an underlying fraction of positions having already accepted destabilizing amino-acids, then the probability of destabilizing mutational events is in turn expected to directly depend on the current stability of the protein.

	In addition, if empirical evidence for a negative correlation of $\omega$ with $\Ne$ is still not totally convincing, another empirical correlation is known to be much more robust, namely, that of $\omega$ with expression level or protein abundance \cite{Duret2000, Rocha2004, Wang2011, Song2017}. Indeed, expression level is one of the best predictors of $\omega$, with highly expressed proteins typically having lower $\omega$ values. Of note, the slope of the response of $\omega$ to changes in expression level is relatively weak \cite{Drummond2005}, although clearly significant. Importantly, many of the theoretical models of protein-coding sequence evolution that have been invoked to explain this correlation between $\omega$ and expression level are such that the response of $\omega$ to changes in expression level or in $\Ne$ should be very similar. Conversely, under strict equimutability, most of these models would predict that $\omega$ should be essentially independent, not just of $\Ne$, but also of expression level, thus at odds with empirical observations.

	Overall, there is no doubt that the impact of changes in $\Ne$ (or in expression level) on $\omega$ is at best weak, making empirical estimation more difficult.
	Even if weak, however, an exact theoretical understanding and quantification of how the slope of this relation depends on the underlying map between genotype, phenotype and fitness, would be useful.
	Ultimately, relating proteins structural parameters to the $\omega$-$\Ne$ relationship reveal an intersection between disparate empirical dataset,  coming from protein thermodynamic on one side and comparative genomic on the other side.

	Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium. However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\omega$ due to adaptive selection, and subsequently a decrease in $\omega$ due to stronger purifying selection in the long term \cite{Jones2016}.
	Studying only equilibrium properties can thus be misleading. For that reason, the dynamical response of $\omega$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized.
	Dynamical properties of the $\omega$ response to changes of $\Ne$ are of theoretical interest, but are also empirically relevant such that if overlook they could thwart the relation between theoretical expectation and empirical estimates.

	% Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\omega$ relationship to $\Ne$ is unknown.
	The goal of this study is to characterize the dynamical and equilibrium response of $\omega$ to changes in $\Ne$, and relate this response to structural parameters of the model.
	To this end, we develop a general mathematical approach for deriving the $\omega$ relationship to $\Ne$ (and expression level) in the context of a given genotype-phenotype-fitness map.
	Based on these results, we discuss the consequences of the assumption that protein stability is a proxy of fitness, and the signatures persisting in DNA sequences observable under such assumption.
	In the light of previously published empirical estimates from protein thermodynamic, and from comparative genomics, we discuss the articulation between empirical data and our mechanistic model.
	We also discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the slope of the relation between $\omega$ and $\Ne$.

@@

	\subsection*{Models of evolution}

	The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$ (i.e. such that the coding positions of the sequence contribute additively to the trait), under directional selection specified by a decreasing and log-concave fitness function $f(x)$. As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability (See figure \ref{fig:NeChangeInfluence}). This model is inspired from previous work \cite{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations. 
	%Throughout our derivation, we present the results both for the general case and for the more specific example.

	In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
	Technically, the free energy of the folded are unfolded conformations are computed based on the $3$D conformations of the protein, and using a statistical potential.
	In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.

	We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighbouring residues.
	Instead, each site contributes independently to $\deltaG$. 
	With such approximation, for each site of the sequence only one amino-acid is stabilizing the protein. 
	All $19$ other amino-acids are equally destabilizing, and each site contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ per destabilizing amino-acids.
	If all amino-acids of the sequence are stabilizing, $\deltaG$ is optimal and equal to $ \alpha < 0$. 
	In this model, the most succinct phenotype of a given genotype (i.e. sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq x \leq 1$. Thus, $\deltaG$ is a linear function of $x$:
	\begin{align}
		\deltaG (x) = \alpha + \Nsite \gamma x,
	\end{align}
	where $\Nsite$ is the number of sites in the sequence.

	For a given $\deltaG$, thermodynamic equations allows to derive the proportion of protein in folded conformation in the cytoplasm, which is assumed to be a proxy for fitness.
	This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm (supp information).
	Analytically, this fitness by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
	\begin{align}
		f(x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma x)}},\\
		\Rightarrow f(x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma x)}, 
	\end{align}
	where $\alpha$ and $\gamma$ are defined as above, and the parameter $\beta$ is $1.686$ mol/kcal at room temperature.

	\subsection*{Susceptibility of $\omega$ to changes in $\Ne$. Analytical approximation}

	In this section we present an analytical approximate solution for the response of equilibrium $\omega$ after a change in $\Ne$ (in log space). We call this reponse the susceptibility of $\omega$ to changes in $\Ne$, and denote it as $\chi$:
	\begin{align}
		\chi = \frac{ \der \omega}{\der \ln (\Ne)}
	\end{align}
	Broadly, deriving $\chi$ is done in two steps.
	The first step is to determine the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
	Subsequently, differential calculus on the response of the equilibrium phenotype to a change in $\Ne$ allows to ultimately derive an equation for $\chi$.
	The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case where the fitness is proportional to the total amount of folded protein in the specific case of the biophysical model.
	All development are available in Supplementary Materials in the most general case.

	First, from a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of destabilizing amino-acid, or do nothing if the mutation is between both destabilizing amino-acids.
	To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable.
	Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
	\begin{gather}
		\begin{cases}
		\dx &\text{ with probability } 1-x, \\
		0 &\text{ with probability } \frac{18 x }{19}, \\
		-\dx &\text{ with probability } \frac{x}{19}.\\
		\end{cases} \label{eq:proba}
	\end{gather}
	In the extreme case of an optimal phenotype ($x = 0$), only destabilization mutations are proposed.
	Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between destabilizing amino-acids (effect $0$) is proportional to $x$. 
	Or in other words, the mutation bias is proportional to $(1-x)/x$, which this is fundamentally a combinatorial effect, due to the number of mutational opportunities available in either direction.

	Second, we need to determine the strength of selection acting on mutations.
	Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
	\begin{align}
		s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(x) }{\partial x} \label{eq:s} \\
		\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma x)}. \label{eq:s-unfolded}
	\end{align}
	Conversely, stabilizing mutations will be under positive selection with opposite sign but same absolute value.
	It is important to realize that the selective effect is dependent on $x$, and that because fitnesses are log-concave, the absolute value of $s$ increases with regards to $x$.

	Based on these expressions for mutational and selective pressure, one can study the evolutionary process unfurls.
	Starting from an optimal sequence, mostly destabilizing mutations will occur, which will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, at which point the protein will reach a point of equilibrium called marginal stability.
	Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
	Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average \cite{Goldstein2013}.
	Formally, and after simplification, the equilibrium phenotype denoted $x\eq$ is given by the equation:
	\begin{align}
		\ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(x\eq) }{\partial {x\eq}} \text{ in the general case,} \\
		\Rightarrow \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)} \text{,} \label{eq:equilibrium}
	\end{align}
	in the more specific case of the biophysical model. This equation essentially expresses the mutation-selection equilibrium: the left-hand side of the equation is the log of the mutation bias at $x$, while the right-hand side is simply $4 \Ne s$.

	This equation cannot be solved explicitly for $x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $x\eq$ is given in figure \ref{fig:NeChangeInfluence}B.
	Mutation bias decreases weakly with $x$ (blue curve on figure) while the strength of selection increases sharply with $x$.
	The equilibrium phenotype is obtain at their intersection (eq \ref{eq:equilibrium}).
	An increase in $\Ne$ leads to shifting the selective response upward, which then results in a leftward shift of the equilbrium phenotype. 
	The amount of change on $x\eq$, however, is very small if the selection coefficient curve is very steep around the equilibrium set point.
	Of note, the phenotype is not equimutable, meaning the mutational bias is not constant, but nearly so since the derivative of the mutational bias is greatly lower than the derivative of the selection coefficient around the equilibrium point.

	\begin{figure*}[htb!]
		\begin{mdframed}
			\centering
			\includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
			\caption{
				\textbf{Theoretical model.}
				Panel A. Illustration of the Genotype-phenotype-fitness map. The phenotype ($x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of destabilizing sites in the sequence. The fitness is a decreasing log-concave function of the phenotype.
				Panel B. Illustration of the $\omega$ susceptibility after a change in $\Ne$. The equilibrium $x\eq$ is determined by the equation $\ln(\frac{1 - x\eq}{x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)}$. The right-hand side of the equation increases exponentially with $x$ where $\Nsite \gamma$ is the exponential growth rate. Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only changes subtly $x\eq$ (x-axis of the crossing with the blue solid line). On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) involve a stronger response of $x\eq$. Moreover, changes in $x\eq$ reflects the changes in $\omega$ since both are related by a monotonous function (equation \ref{eq:dnds}). The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
			}
			\label{fig:NeChangeInfluence}
		\end{mdframed}
	\end{figure*}

	The argument thus far explains how $x\eq$ varies with $\Ne$.
	To capture how $\omega$ varies with $\Ne$, we also need to obtain an expression for $\omega$ as a function of $x\eq$. 
	At equilibrium we can derive the expected substitution rate of mutations changing the amino-acid sequence, and by normalizying by the substitution rate of neutral mutations, $\omega$ simply approximates to:
	\begin{gather}
		\omega \simeq x\eq \label{eq:dnds}
	\end{gather}
	This simple approximation is due to substitutions at destabilizing sites (proportion $x\eq$) to one of the other destabilizing amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (see equation \ref{eq:proba}).
